Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-12 paper-new/intro.tex
tweaks to intro and syntax sections

In this paper, we develop an adequate denotational semantics that satisfies
graduality and soundness of the equational theory of cast calculi using the
techniques of SGDT.

@@

\begin{enumerate}
\item First, we give a simple concrete term semantics where we show
  how to model the dynamic type as a solution to a guarded domain equation
\item Next, we identify where prior work on classical domain theoretic
  semantics of gradual typing breaks down when using guarded semantics
  of recursive types.
\item We develop a key new concept of \emph{syntactic perturbations},
  which allow us to recover enough extensional reasoning to model the
  graduality property compositionally
\item We combine this insight together with an abstract categorical
  model of gradual typing using reflexive graph categories and
  call-by-push-value to give a compositional construction of our
  denotational model.
\item We prove that the resulting denotational model provides a
  well-behaved semantics as defined above by proving \emph{adequacy},
  respect for an equational theory and the graduality property.
\end{enumerate}

The paper is laid out as follows:
\begin{enumerate}
\item In Section\ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section\ref{sec:concrete-term-model} we develop a first
  denotational semantics in synthetic guarded domain theory that
  satisfies adequacy and respects the equational theory, but does not
  validate graduality. We use this to introduce some of our main
  technical tools: modeling recursive types in guarded type theory and
  modeling effects using call-by-push-value.
\item In Section\ref{sec:towards-relational-model} we show where the
  New-Licata classical domain theoretic approach fails to adapt
  cleanly to the guarded setting and explore the difficulties of
  proving graduality in an intensional model.
\item In Section\ref{sec:concrete-relational-model} we describe the resulting
  model in detail. We then prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  N : \nat$ has the expected semantics, that $M$ errors or $M$ and $N$ have the
  same extensional behavior.
\item In Section\ref{sec:discussion} we discuss prior work on proving
  graduality, our partial mechanization of these results and discuss
  future directions for denotational semantics of gradual typing.
\end{enumerate}

>>>>

In this paper, we develop an adequate denotational semantics that satisfies
graduality and soundness of the equational theory of cast calculi using
synthetic guarded domain theory.

@@

\begin{enumerate}
\item First, we give a simple concrete term semantics where we show
  how to model the dynamic type as a solution to a guarded domain equation.
\item Next, we identify where prior work on classical domain theoretic
  semantics of gradual typing breaks down when using guarded semantics
  of recursive types.
\item We develop a key new concept of \emph{syntactic perturbations},
  which allow us to recover enough extensional reasoning to model the
  graduality property compositionally.
\item We combine this insight together with an abstract categorical
  model of gradual typing using reflexive graph categories and
  call-by-push-value to give a compositional construction of our
  denotational model.
\item We prove that the resulting denotational model provides a
  well-behaved semantics as defined above by proving \emph{adequacy},
  respect for an equational theory and the graduality property.
\end{enumerate}

The paper is laid out as follows:
\begin{enumerate}
\item In Section \ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section \ref{sec:concrete-term-model} we develop a first
  denotational semantics in synthetic guarded domain theory that
  satisfies adequacy and respects the equational theory, but does not
  validate graduality. We use this to introduce some of our main
  technical tools: modeling recursive types in guarded type theory and
  modeling effects using call-by-push-value.
\item In Section \ref{sec:towards-relational-model} we show where the
  New-Licata classical domain theoretic approach fails to adapt
  cleanly to the guarded setting and explore the difficulties of
  proving graduality in an intensional model.
\item In Section \ref{sec:concrete-relational-model} we describe the resulting
  model in detail. We then prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  N : \nat$ has the expected semantics, i.e., that $M$ errors or $M$ and $N$ have the
  same extensional behavior.
\item In Section \ref{sec:discussion} we discuss prior work on proving
  graduality, our partial mechanization of these results and discuss
  future directions for denotational semantics of gradual typing.
\end{enumerate}
